---
title: "Extreme Heat Indicators"
output: html_document
date: "2024-12-23"
author: "Climate Services, CA Governor's Office of Land Use and Climate Innovation
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extreme Heat Analysis for Vulnerable Communities Platform

## Data Description

This code utilizes downscaled climate data available from the Cal-Adapt Analytics Engine to generate an indicator of extreme heat for the Vulnerable Communities Platform (CA Governor's Office of Land Use and Climate Innovation). Specifically, we use downscaled tmaxdata from ACCESS-CM2 r1i1p1f1 for SSP370 for two different time scales (2015-2044 and 2045-2074).

The data are available from the following links:

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/ssp370/tasmax/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/historical/tasmax/

## Frequency of days over 100 degrees F for 2045-2074 (VCP 2050 timescale)

Counts the number of days over June, July, and August of 2045-2074 where daily temperatures exceed 100 degrees Farenheit at each location in the grid.

```{r}
library(ncdf4)

# Path to NetCDF file
ncfile <- "tasmax.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Open the NetCDF file
nc <- nc_open(ncfile)

# Extract temperature data and coordinates
lon <- ncvar_get(nc, "lon")
lat <- ncvar_get(nc, "lat")
time <- ncvar_get(nc, "time")

# Set up the time period for June, July, and August
time_origin <- as.Date("1900-01-01")
dates <- time_origin + time
subset_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Create an empty array to store the exceedance count
exceedance_count <- array(0, dim = c(length(lon), length(lat)))

# Loop through the subset of time indices (JJA)
for (i in subset_indices) {
  # Read the temperature data for the current time index
  temperature_chunk <- ncvar_get(nc, "tasmax", start = c(1, 1, i), count = c(length(lon), length(lat), 1))
  
  # Threshold for 100°F in Kelvin
  threshold_fahrenheit_kelvin <- 310.928
  
  # Calculate the exceedances for the current chunk
  exceedance_chunk <- temperature_chunk > threshold_fahrenheit_kelvin
  
  # Add the exceedance counts to the total count
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Define the dimensions for the new NetCDF file
lon_dim <- ncdim_def("lon", "degrees_east", lon)
lat_dim <- ncdim_def("lat", "degrees_north", lat)
time_dim <- ncdim_def("time", "days", 1)

# Define the exceedance_count variable
exceedance_var <- ncvar_def("exceedance_count", "count", list(lon_dim, lat_dim, time_dim), 
                            missval = NA, prec = "double")

# Create a new NetCDF file to save the exceedance count
new_ncfile <- "exceedance_100F_2045-2074.nc"
nc_new <- nc_create(new_ncfile, list(exceedance_var))

# Write the exceedance count data into the new NetCDF file
ncvar_put(nc_new, "exceedance_count", exceedance_count)

# Add coordinate variables (latitude, longitude) and time to the new NetCDF file
ncvar_put(nc_new, "lon", lon)
ncvar_put(nc_new, "lat", lat)
ncvar_put(nc_new, "time", 1)  # Assuming single time dimension (for JJA)

# Close the new NetCDF file
nc_close(nc_new)

# Close the original NetCDF file
nc_close(nc)
```

## Frequency of days over 100 degrees F for 2015-2044 (VCP current timescale) 

Counts the number of days over June, July, and August of 2015-2044 where daily temperatures exceed 100 degrees Farenheit at each location in the grid.

```{r}
library(ncdf4)

# Path to NetCDF file
ncfile <- "tasmax.ACCESS-CM2.ssp370.r1i1p1f1.2015-2044.LOCA2_CA_hybrid_v20230115.nc"

# Open the NetCDF file
nc <- nc_open(ncfile)

# Extract temperature data and coordinates
lon <- ncvar_get(nc, "lon")
lat <- ncvar_get(nc, "lat")
time <- ncvar_get(nc, "time")

# Set up the time period for June, July, and August
time_origin <- as.Date("1900-01-01")
dates <- time_origin + time
subset_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Create an empty array to store the exceedance count
exceedance_count <- array(0, dim = c(length(lon), length(lat)))

# Loop through the subset of time indices (JJA)
for (i in subset_indices) {
  # Read the temperature data for the current time index
  temperature_chunk <- ncvar_get(nc, "tasmax", start = c(1, 1, i), count = c(length(lon), length(lat), 1))
  
  # Threshold for 100°F in Kelvin
  threshold_fahrenheit_kelvin <- 310.928
  
  # Calculate the exceedances for the current chunk
  exceedance_chunk <- temperature_chunk > threshold_fahrenheit_kelvin
  
  # Add the exceedance counts to the total count
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Define the dimensions for the new NetCDF file
lon_dim <- ncdim_def("lon", "degrees_east", lon)
lat_dim <- ncdim_def("lat", "degrees_north", lat)
time_dim <- ncdim_def("time", "days", 1)

# Define the exceedance_count variable
exceedance_var <- ncvar_def("exceedance_count", "count", list(lon_dim, lat_dim, time_dim), 
                            missval = NA, prec = "double")

# Create a new NetCDF file to save the exceedance count
new_ncfile <- "exceedance_100F_2015-2044.nc"
nc_new <- nc_create(new_ncfile, list(exceedance_var))

# Write the exceedance count data into the new NetCDF file
ncvar_put(nc_new, "exceedance_count", exceedance_count)

# Add coordinate variables (latitude, longitude) and time to the new NetCDF file
ncvar_put(nc_new, "lon", lon)
ncvar_put(nc_new, "lat", lat)
ncvar_put(nc_new, "time", 1)  # Assuming single time dimension (for JJA)

# Close the new NetCDF file
nc_close(nc_new)

# Close the original NetCDF file
nc_close(nc)
```
