---
title: "Extreme Heat Indicators"
output: html_document
date: "2024-12-23"
author: "Climate Services, CA Governor's Office of Land Use and Climate Innovation
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extreme Heat Analysis for Vulnerable Communities Platform

## Data Description

This code utilizes downscaled climate data available from the Cal-Adapt Analytics Engine to generate indicators of extreme heat for the Vulnerable Communities Platform (CA Governor's Office of Land Use and Climate Innovation). Specifically, we use downscaled tmin data from ACCESS-CM2 r1i1p1f1 for SSP370 for two different time scales (2015-2044 and 2045-2074) and modeled historical baseline data sets for tmin and tmax from 1950-2014.

The data are available from the following links:

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/historical/tasmax/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/historical/tasmin/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/ssp370/tasmin/ 


## Frequency of extreme heat nights (as defined by the 98th percentile) for 2045-2074 (VCP 2050 timescale)

Counts the number of extreme heat nights over June, July, and August of 2045-2074 for each location in the grid, where extreme heat nights are defined by an exceedance of the 98th historical (1950-2014) percentile of minimum temperature. Only locations that have a historical 98th percentile of tmax >= 90 deg F are considered.

```{r}
library(terra)
library(ncdf4)

# File paths
historical_tasmax_ncfile <- "tasmax.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
historical_tasmin_ncfile <- "tasmin.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
future_tasmin_ncfile <- "tasmin.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Load historical tasmax and tasmin
r_tasmax_hist <- rast(historical_tasmax_ncfile)
r_tasmin_hist <- rast(historical_tasmin_ncfile)

# Open files
nc_hist_max <- nc_open(historical_tasmax_ncfile)

# Identify JJA months
time_origin <- as.Date("1900-01-01")
time_vals <- ncvar_get(nc_hist_max, "time")
dates <- time_origin + time_vals
jja_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Compute 98th percentile of tasmax for JJA (1950-2014)
p90_tasmax <- app(r_tasmax_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Compute 98th percentile of tasmin for JJA (1950-2014)
p98_tasmin <- app(r_tasmin_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Filter locations where 98th percentile tasmax >= 90°F (305.4K)
valid_locations <- p90_tasmax >= 305.4

# Load future tasmin
r_tasmin_future <- rast(future_tasmin_ncfile)

# Open files
nc_fut_min <- nc_open(future_tasmin_ncfile)

# Get future JJA indices
time_vals_future <- ncvar_get(nc_fut_min, "time")
print(time_vals_future)
dates_future <- as.Date(time_vals_future, origin = "1900-01-01")
print(dates_future)
jja_indices_future <- which(format(dates_future, "%m") %in% c("06", "07", "08"))
print(jja_indices_future)

# Exceedance count
exceedance_count <- deepcopy(r_tasmin_future[[1]])
values(exceedance_count) <- 0

# Process in chunks
for (i in jja_indices_future) {
  chunk <- r_tasmin_future[[i]]
  exceedance_chunk <- (chunk > p98_tasmin) & valid_locations
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Save exceedance count
out_ncfile <- "exceedance_98th_percentile_filtered_tmin_2045-2074.nc"
writeRaster(exceedance_count, out_ncfile, overwrite = TRUE, filetype="netcdf")
```

## Frequency of extreme heat nights (as defined by the 98th percentile) for 2015-2044 (VCP current timescale)

Counts the number of extreme heat nights over June, July, and August of 2015-2044 for each location in the grid, where extreme heat nights are defined by an exceedance of the 98th historical (1950-2014) percentile of minimum temperature. Only locations that have a historical 90th percentile of tmax >= 90 deg F are considered.

```{r}
library(terra)
library(ncdf4)

# File paths
historical_tasmax_ncfile <- "tasmax.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
historical_tasmin_ncfile <- "tasmin.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
future_tasmin_ncfile <- "tasmin.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Load historical tasmax and tasmin
r_tasmax_hist <- rast(historical_tasmax_ncfile)
r_tasmin_hist <- rast(historical_tasmin_ncfile)

# Open files
nc_hist_max <- nc_open(historical_tasmax_ncfile)

# Identify JJA months
time_origin <- as.Date("1900-01-01")
time_vals <- ncvar_get(nc_hist_max, "time")
dates <- time_origin + time_vals
jja_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Compute 98th percentile of tasmax for JJA (1950-2014)
p90_tasmax <- app(r_tasmax_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Compute 98th percentile of tasmin for JJA (1950-2014)
p98_tasmin <- app(r_tasmin_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Filter locations where 98th percentile tasmax >= 90°F (305.4K)
valid_locations <- p90_tasmax >= 305.4

# File path
future_tasmin_ncfile <- "tasmin.ACCESS-CM2.ssp370.r1i1p1f1.2015-2044.LOCA2_CA_hybrid_v20230115.nc"

# Load future tasmin
r_tasmin_future <- rast(future_tasmin_ncfile)

# Open NC file
nc_future_min <- nc_open(future_tasmin_ncfile)

# Get future JJA indices
time_vals_future <- ncvar_get(nc_future_min, "time")
print(time_vals_future)
dates_future <- as.Date(time_vals_future, origin = "1900-01-01")
print(dates_future)
jja_indices_future <- which(format(dates_future, "%m") %in% c("06", "07", "08"))
print(jja_indices_future)

#Exceedance count
exceedance_count <- deepcopy(r_tasmin_future[[1]])
values(exceedance_count) <- 0

# Process in chunks
for (i in jja_indices_future) {
  chunk <- r_tasmin_future[[i]]
  exceedance_chunk <- (chunk > p98_tasmin) & valid_locations
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Save exceedance count
out_ncfile <- "exceedance_98th_percentile_filtered_tmin_2015-2044.nc"
writeRaster(exceedance_count, out_ncfile, overwrite = TRUE, filetype="netcdf")

```
