---
title: "Extreme Heat Indicators"
output: html_document
date: "2024-12-23"
author: "Climate Services, CA Governor's Office of Land Use and Climate Innovation
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Extreme Heat Analysis for Vulnerable Communities Platform

## Data Description

This code utilizes downscaled climate data available from the Cal-Adapt Analytics Engine to generate indicators of extreme heat for the Vulnerable Communities Platform (CA Governor's Office of Land Use and Climate Innovation). Specifically, we use downscaled tmax and tmin data from ACCESS-CM2 r1i1p1f1 for SSP370 for two different time scales (2015-2044 and 2045-2074) and modeled historical baseline data sets from 1950-2014.

The data are available from the following links:

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/ssp370/tasmax/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/historical/tasmax/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/historical/tasmin/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/ssp370/tasmin/ 

## Frequency of days over 100 degrees F for 2045-2074 (VCP 2050 timescale)

Counts the number of days over June, July, and August of 2045-2074 where daily temperatures exceed 100 degrees Farenheit at each location in the grid.

```{r}
library(ncdf4)

# Path to NetCDF file
ncfile <- "tasmax.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Open the NetCDF file
nc <- nc_open(ncfile)

# Extract temperature data and coordinates
lon <- ncvar_get(nc, "lon")
lat <- ncvar_get(nc, "lat")
time <- ncvar_get(nc, "time")

# Set up the time period for June, July, and August
time_origin <- as.Date("1900-01-01")
dates <- time_origin + time
subset_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Create an empty array to store the exceedance count
exceedance_count <- array(0, dim = c(length(lon), length(lat)))

# Loop through the subset of time indices (JJA)
for (i in subset_indices) {
  # Read the temperature data for the current time index
  temperature_chunk <- ncvar_get(nc, "tasmax", start = c(1, 1, i), count = c(length(lon), length(lat), 1))
  
  # Threshold for 100째F in Kelvin
  threshold_fahrenheit_kelvin <- 310.928
  
  # Calculate the exceedances for the current chunk
  exceedance_chunk <- temperature_chunk > threshold_fahrenheit_kelvin
  
  # Add the exceedance counts to the total count
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Define the dimensions for the new NetCDF file
lon_dim <- ncdim_def("lon", "degrees_east", lon)
lat_dim <- ncdim_def("lat", "degrees_north", lat)
time_dim <- ncdim_def("time", "days", 1)

# Define the exceedance_count variable
exceedance_var <- ncvar_def("exceedance_count", "count", list(lon_dim, lat_dim, time_dim), 
                            missval = NA, prec = "double")

# Create a new NetCDF file to save the exceedance count
new_ncfile <- "exceedance_100F_2045-2074.nc"
nc_new <- nc_create(new_ncfile, list(exceedance_var))

# Write the exceedance count data into the new NetCDF file
ncvar_put(nc_new, "exceedance_count", exceedance_count)

# Add coordinate variables (latitude, longitude) and time to the new NetCDF file
ncvar_put(nc_new, "lon", lon)
ncvar_put(nc_new, "lat", lat)
ncvar_put(nc_new, "time", 1)  # Assuming single time dimension (for JJA)

# Close the new NetCDF file
nc_close(nc_new)

# Close the original NetCDF file
nc_close(nc)
```

## Frequency of days over 100 degrees F for 2015-2044 (VCP current timescale) 

Counts the number of days over June, July, and August of 2015-2044 where daily temperatures exceed 100 degrees Farenheit at each location in the grid.

```{r}
library(ncdf4)

# Path to NetCDF file
ncfile <- "tasmax.ACCESS-CM2.ssp370.r1i1p1f1.2015-2044.LOCA2_CA_hybrid_v20230115.nc"

# Open the NetCDF file
nc <- nc_open(ncfile)

# Extract temperature data and coordinates
lon <- ncvar_get(nc, "lon")
lat <- ncvar_get(nc, "lat")
time <- ncvar_get(nc, "time")

# Set up the time period for June, July, and August
time_origin <- as.Date("1900-01-01")
dates <- time_origin + time
subset_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Create an empty array to store the exceedance count
exceedance_count <- array(0, dim = c(length(lon), length(lat)))

# Loop through the subset of time indices (JJA)
for (i in subset_indices) {
  # Read the temperature data for the current time index
  temperature_chunk <- ncvar_get(nc, "tasmax", start = c(1, 1, i), count = c(length(lon), length(lat), 1))
  
  # Threshold for 100째F in Kelvin
  threshold_fahrenheit_kelvin <- 310.928
  
  # Calculate the exceedances for the current chunk
  exceedance_chunk <- temperature_chunk > threshold_fahrenheit_kelvin
  
  # Add the exceedance counts to the total count
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Define the dimensions for the new NetCDF file
lon_dim <- ncdim_def("lon", "degrees_east", lon)
lat_dim <- ncdim_def("lat", "degrees_north", lat)
time_dim <- ncdim_def("time", "days", 1)

# Define the exceedance_count variable
exceedance_var <- ncvar_def("exceedance_count", "count", list(lon_dim, lat_dim, time_dim), 
                            missval = NA, prec = "double")

# Create a new NetCDF file to save the exceedance count
new_ncfile <- "exceedance_100F_2015-2044.nc"
nc_new <- nc_create(new_ncfile, list(exceedance_var))

# Write the exceedance count data into the new NetCDF file
ncvar_put(nc_new, "exceedance_count", exceedance_count)

# Add coordinate variables (latitude, longitude) and time to the new NetCDF file
ncvar_put(nc_new, "lon", lon)
ncvar_put(nc_new, "lat", lat)
ncvar_put(nc_new, "time", 1)  # Assuming single time dimension (for JJA)

# Close the new NetCDF file
nc_close(nc_new)

# Close the original NetCDF file
nc_close(nc)
```

## Frequency of extreme heat nights (as defined by the 98th percentile) for 2045-2074 (VCP 2050 timescale)

Counts the number of extreme heat nights over June, July, and August of 2045-2074 for each location in the grid, where extreme heat nights are defined by an exceedance of the 98th historical (1950-2014) percentile of minimum temperature. Only locations that have a historical 98th percentile of tmax >= 90 deg F are considered.

```{r}
library(terra)
library(ncdf4)

# File paths
historical_tasmax_ncfile <- "tasmax.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
historical_tasmin_ncfile <- "tasmin.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
future_tasmin_ncfile <- "tasmin.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Load historical tasmax and tasmin
r_tasmax_hist <- rast(historical_tasmax_ncfile)
r_tasmin_hist <- rast(historical_tasmin_ncfile)

# Open files
nc_hist_max <- nc_open(historical_tasmax_ncfile)

# Identify JJA months
time_origin <- as.Date("1900-01-01")
time_vals <- ncvar_get(nc_hist_max, "time")
dates <- time_origin + time_vals
jja_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Compute 98th percentile of tasmax for JJA (1950-2014)
p90_tasmax <- app(r_tasmax_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Compute 98th percentile of tasmin for JJA (1950-2014)
p98_tasmin <- app(r_tasmin_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Filter locations where 98th percentile tasmax >= 90째F (305.4K)
valid_locations <- p90_tasmax >= 305.4

# Load future tasmin
r_tasmin_future <- rast(future_tasmin_ncfile)

# Open files
nc_fut_min <- nc_open(future_tasmin_ncfile)

# Get future JJA indices
time_vals_future <- ncvar_get(nc_fut_min, "time")
print(time_vals_future)
dates_future <- as.Date(time_vals_future, origin = "1900-01-01")
print(dates_future)
jja_indices_future <- which(format(dates_future, "%m") %in% c("06", "07", "08"))
print(jja_indices_future)

# Exceedance count
exceedance_count <- deepcopy(r_tasmin_future[[1]])
values(exceedance_count) <- 0

# Process in chunks
for (i in jja_indices_future) {
  chunk <- r_tasmin_future[[i]]
  exceedance_chunk <- (chunk > p98_tasmin) & valid_locations
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Save exceedance count
out_ncfile <- "exceedance_98th_percentile_filtered_tmin_2045-2074.nc"
writeRaster(exceedance_count, out_ncfile, overwrite = TRUE, filetype="netcdf")
```

## Frequency of extreme heat nights (as defined by the 98th percentile) for 2015-2044 (VCP current timescale)

Counts the number of extreme heat nights over June, July, and August of 2015-2044 for each location in the grid, where extreme heat nights are defined by an exceedance of the 98th historical (1950-2014) percentile of minimum temperature. Only locations that have a historical 90th percentile of tmax >= 90 deg F are considered.

```{r}
library(terra)
library(ncdf4)

# File paths
historical_tasmax_ncfile <- "tasmax.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
historical_tasmin_ncfile <- "tasmin.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
future_tasmin_ncfile <- "tasmin.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Load historical tasmax and tasmin
r_tasmax_hist <- rast(historical_tasmax_ncfile)
r_tasmin_hist <- rast(historical_tasmin_ncfile)

# Open files
nc_hist_max <- nc_open(historical_tasmax_ncfile)

# Identify JJA months
time_origin <- as.Date("1900-01-01")
time_vals <- ncvar_get(nc_hist_max, "time")
dates <- time_origin + time_vals
jja_indices <- which(format(dates, "%m") %in% c("06", "07", "08"))

# Compute 98th percentile of tasmax for JJA (1950-2014)
p90_tasmax <- app(r_tasmax_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Compute 98th percentile of tasmin for JJA (1950-2014)
p98_tasmin <- app(r_tasmin_hist[[jja_indices]], fun = function(x) quantile(x, 0.98, na.rm = TRUE))

# Filter locations where 98th percentile tasmax >= 90째F (305.4K)
valid_locations <- p90_tasmax >= 305.4

# File path
future_tasmin_ncfile <- "tasmin.ACCESS-CM2.ssp370.r1i1p1f1.2015-2044.LOCA2_CA_hybrid_v20230115.nc"

# Load future tasmin
r_tasmin_future <- rast(future_tasmin_ncfile)

# Open NC file
nc_future_min <- nc_open(future_tasmin_ncfile)

# Get future JJA indices
time_vals_future <- ncvar_get(nc_future_min, "time")
print(time_vals_future)
dates_future <- as.Date(time_vals_future, origin = "1900-01-01")
print(dates_future)
jja_indices_future <- which(format(dates_future, "%m") %in% c("06", "07", "08"))
print(jja_indices_future)

#Exceedance count
exceedance_count <- deepcopy(r_tasmin_future[[1]])
values(exceedance_count) <- 0

# Process in chunks
for (i in jja_indices_future) {
  chunk <- r_tasmin_future[[i]]
  exceedance_chunk <- (chunk > p98_tasmin) & valid_locations
  exceedance_count <- exceedance_count + exceedance_chunk
}

# Save exceedance count
out_ncfile <- "exceedance_98th_percentile_filtered_tmin_2015-2044.nc"
writeRaster(exceedance_count, out_ncfile, overwrite = TRUE, filetype="netcdf")

```
