---
title: "Wildfire Indicators"
output: html_document
date: "2024-12-23"
author: "Climate Services, CA Governor's Office of Land Use and Climate Innovation
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Wildfire Analysis for Vulnerable Communities Platform

## Data Description

This code utilizes downscaled climate data available from the Cal-Adapt Analytics Engine to generate an indicator for Wildfire for the Vulnerable Communities Platform (CA Governor's Office of Land Use and Climate Innovation). Specifically, we use downscaled tmax data from ACCESS-CM2 r1i1p1f1 for SSP370 for two different time scales (2015-2044 and 2045-2074) and modeled historical baseline data from 1950-2014.

The data are available from the following links:

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/ssp370/tasmax/

https://cadcat.s3.amazonaws.com/index.html#loca2/aaa-ca-hybrid/ACCESS-CM2/0p0625deg/r1i1p1f1/historical/tasmax/

## Change in average Sept, Oct, Nov temperature (2050 timescale)

Calculates the change in average maximum temperature observations in September, October, November between 2045-2074 and the historical baseline (1950-2014) at each location in the grid.

```{r}
library(ncdf4)

# Function to calculate mean temperature for a given NetCDF file and period
temp_mean_SON <- function(ncfile, start_year, end_year) {
  nc <- nc_open(ncfile)
  
  time <- ncvar_get(nc, "time")
  time_origin <- as.Date("1900-01-01")
  dates <- time_origin + time
  
  # Extract indices for September-November
  subset_indices <- which(format(dates, "%m") %in% c("09", "10", "11") & 
                            format(dates, "%Y") %in% as.character(start_year:end_year))
  
  # Get spatial dimensions
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  
  # Initialize sum and count arrays
  temp_sum <- array(0, dim = c(length(lon), length(lat)))
  count <- array(0, dim = c(length(lon), length(lat)))
  
  for (i in subset_indices) {
    temp_chunk <- ncvar_get(nc, "tasmax", start = c(1, 1, i), count = c(length(lon), length(lat), 1))
    temp_sum <- temp_sum + temp_chunk
    count <- count + 1
  }
  
  nc_close(nc)
  
  return(list(mean_temp = temp_sum / count, lon = lon, lat = lat))  # Return temp, lon, and lat
}

# Define NetCDF files
historical_ncfile <- "tasmax.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
future_ncfile <- "tasmax.ACCESS-CM2.ssp370.r1i1p1f1.2045-2074.LOCA2_CA_hybrid_v20230115.nc"

# Compute historical and future mean temperature (still in Kelvin)
historical_result <- temp_mean_SON(historical_ncfile, 1950, 2014)
future_result <- temp_mean_SON(future_ncfile, 2045, 2074)

# Convert mean temperatures from Kelvin to Fahrenheit
temp_mean_historical <- historical_result$mean_temp * 1.8 - 459.67
temp_mean_future <- future_result$mean_temp * 1.8 - 459.67

# Extract lon and lat
lon <- historical_result$lon
lat <- historical_result$lat

# Compute change in mean temperature (already in Fahrenheit)
delta_temp <- temp_mean_future - temp_mean_historical  

# Define NetCDF output file
new_ncfile <- "change_mean_temp_SON_2045_2074_Fahrenheit.nc"
lon_dim <- ncdim_def("lon", "degrees_east", lon)
lat_dim <- ncdim_def("lat", "degrees_north", lat)
delta_var <- ncvar_def("delta_temp", "degrees_F", list(lon_dim, lat_dim), missval = NA, prec = "double")

# Create and save NetCDF file
nc_new <- nc_create(new_ncfile, list(delta_var))
ncvar_put(nc_new, "delta_temp", delta_temp)
ncvar_put(nc_new, "lon", lon)
ncvar_put(nc_new, "lat", lat)
nc_close(nc_new)

```

## Change in average Sept, Oct, Nov temperature (current timescale)

Calculates the change in average maximum temperature observations in September, October, November between 2015-2044 and the historical baseline (1950-2014) at each location in the grid.

```{r}
library(ncdf4)

# Function to calculate mean temperature for a given NetCDF file and period
temp_mean_SON <- function(ncfile, start_year, end_year) {
  nc <- nc_open(ncfile)
  
  time <- ncvar_get(nc, "time")
  time_origin <- as.Date("1900-01-01")
  dates <- time_origin + time
  
  # Extract indices for September-November
  subset_indices <- which(format(dates, "%m") %in% c("09", "10", "11") & 
                            format(dates, "%Y") %in% as.character(start_year:end_year))
  
  # Get spatial dimensions
  lon <- ncvar_get(nc, "lon")
  lat <- ncvar_get(nc, "lat")
  
  # Initialize sum and count arrays
  temp_sum <- array(0, dim = c(length(lon), length(lat)))
  count <- array(0, dim = c(length(lon), length(lat)))
  
  for (i in subset_indices) {
    temp_chunk <- ncvar_get(nc, "tasmax", start = c(1, 1, i), count = c(length(lon), length(lat), 1))
    temp_sum <- temp_sum + temp_chunk
    count <- count + 1
  }
  
  nc_close(nc)
  
  return(list(mean_temp = temp_sum / count, lon = lon, lat = lat))  # Return temp, lon, and lat
}

# Define NetCDF files
historical_ncfile <- "tasmax.ACCESS-CM2.historical.r1i1p1f1.1950-2014.LOCA2_CA_hybrid_v20230115.nc"
future_ncfile <- "tasmax.ACCESS-CM2.ssp370.r1i1p1f1.2015-2044.LOCA2_CA_hybrid_v20230115.nc"

# Compute historical and future mean temperature (still in Kelvin)
historical_result <- temp_mean_SON(historical_ncfile, 1950, 2014)
future_result <- temp_mean_SON(future_ncfile, 2015, 2044)

# Convert mean temperatures from Kelvin to Fahrenheit
temp_mean_historical <- historical_result$mean_temp * 1.8 - 459.67
temp_mean_future <- future_result$mean_temp * 1.8 - 459.67

# Extract lon and lat
lon <- historical_result$lon
lat <- historical_result$lat

# Compute change in mean temperature (already in Fahrenheit)
delta_temp <- temp_mean_future - temp_mean_historical  

# Define NetCDF output file
new_ncfile <- "change_mean_temp_SON_2015_2044_Fahrenheit.nc"
lon_dim <- ncdim_def("lon", "degrees_east", lon)
lat_dim <- ncdim_def("lat", "degrees_north", lat)
delta_var <- ncvar_def("delta_temp", "degrees_F", list(lon_dim, lat_dim), missval = NA, prec = "double")

# Create and save NetCDF file
nc_new <- nc_create(new_ncfile, list(delta_var))
ncvar_put(nc_new, "delta_temp", delta_temp)
ncvar_put(nc_new, "lon", lon)
ncvar_put(nc_new, "lat", lat)
nc_close(nc_new)

```
